---
output: 
  md_document:
    variant: gfm
---
[![Travis build status](https://travis-ci.org/levisc8/ipmr.svg?branch=master)](https://travis-ci.org/levisc8/ipmr)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/levisc8/ipmr?branch=master&svg=true)](https://ci.appveyor.com/project/levisc8/ipmr)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/ipmr)](https://cran.r-project.org/package=ipmr)
[![Codecov test coverage](https://codecov.io/gh/levisc8/ipmr/branch/master/graph/badge.svg)](https://codecov.io/gh/levisc8/ipmr?branch=master)

# ipmr

`ipmr` is a package for implementing Integral Projection Models (IPMs) in _R_. It relies heavily on the mathematical syntax of the models, and does not try to abstract over the process of fitting vital rates. It is still very much a work in progress, and so models implemented using current code may not run in the not-so-distant future as more complicated methods are brought online and tweaks are made. Below is a brief overview of how `ipmr` classifies different model types followed by examples of how to implement those types in this framework. 

## Installation

`ipmr` is not yet on CRAN. You can install the development version with the snippet below:

```{r eval = FALSE}

if(!require('remotes', quietly = TRUE)) {
  install.packages("remotes")
}

remotes::install_github("levisc8/ipmr")

```

## Package scope

Note that this package **will not** help with the process of fitting vital rate models at all! That is a sufficiently different (and vast) question that we decided it was not within the scope of this project. This will only help you turn those regression models into an IPM without shooting yourself in the foot. Thus, everything that follows assumes you have parameterized those vital rate models and are now ready to begin implementing your model. 

Below is a brief overview of the package and some examples of how to implement models with it.

## Model classes

The first step of defining a model in `ipmr` (assuming all parameters have already been estimated) is to initialize the model using `init_ipm()`. This function takes a single argument: `model_class`. The `model_class` defines the basic infrastructure that will be available for subsequent analyses and helps make sure the kernels are correctly implemented from the underlying vital rates. `model_class` should be a character string with at least 3 (but possibly 4) entries separated by underscores (`_`). The possible entries for each position are:


- Position 1: `"simple"`/`"general"`

    + A. **simple**: This describes an IPM with a single continuous state variable and no discrete stages. 

    + B. **general**: This describes and IPM with either more than one continuous state variable, one or more discrete stages, or both of the above. Basically, anything other than an IPM with a single continuous state variable.

- Position 2: `"di"`/`"dd"`

    + A. **di**: This is used to denote a density-independent IPM.

    + B. **dd**: This is used to denote a density-dependent IPM.

- Position 3: `"det"`/`"stoch"`

    + A. **det**: This is used to denote a deterministic IPM. If this is used in the third position of `model_class`, there should not be a fourth entry.

    + B. **stoch**: This is used to denote a stochastic IPM. If this is used in the third position of `model_class`, there should always be a fourth entry. The two possibilities for the fourth are described next.

- Position 4: `"kern"`/`"param"`

    + A. **kern**: This describes an IPM with discretely varying parameters such that their values are known before the model is specified. This is usually the case with models that estimate random year/site effects and for which defining a multivariate joint distribution to sample parameters from is not desirable/needed. These models can be a bit more computationally efficient than the `param` alternative because all kernels can be constructed before the iteration procedure begins, as opposed to requiring reconstruction for every single iteration. 

    + B. **param**: This describes an IPM with parameters that are re-sampled from some distribution at each iteration of the model (usually a multivariate joint distribution). This can be a multivariate normal defined by covarying slopes and intercepts, or posterior distribution from a Bayesian model. All that is required is that the parameters for the distribution are specified and that the function that generates the parameters at each iteration returns named lists that correspond to the parameter names in the model. Jump down to the `"simple_di_stoch_param"` example for some inspiration in writing those.

With the type of model selected, the `model_class` becomes a string and the call to `init_ipm` is composed like so: `init_ipm(model_class = "position1_position_2_position3_position4")`. 

The following possibilities are currently or will become available in `ipmr` (bold text denotes development progress):

- Simple, density independent models: **Completed and ready**

1. `"simple_di_det"`

2. `"simple_di_stoch_kern"`

3. `"simple_di_stoch_param"`

- Simple, density dependent models: **Please be patient**

4. `"simple_dd_det"`

5. `"simple_dd_stoch_kern"`

6. `"simple_dd_stoch_param"`

- General, density independent models: **Completed and ready**

7. `"general_di_det"`

8. `"general_di_stoch_kern"`

9. `"general_di_stoch_param"`

-
General, density dependent models: **Please be patient**

10. `"general_dd_det"`

11. `"general_dd_stoch_kern"`

12. `"general_dd_stoch_param"`

Simple density-independent deterministic, simple kernel-resampled stochastic, and simple parameter resampled stochastic models (`simple_di_det`, `simple_di_stoch_kern`, `simple_di_stoch_param`) are now functional, and are introduced in detail [here](https://levisc8.github.io/ipmr/articles/ipmr-introduction.html). The `general_*` versions of these are also ready, and an introduction to them is available [here](https://levisc8.github.io/ipmr/articles/general-ipms.html). However, expect changes as more complicated methods are implemented! See below for an example of how to implement an IPM in this framework.

## Examples for simple IPM classes

Here is a simple model implemented with `ipmr`. It will use the following set of linear models:

1. Survival (`s`): a generalized linear model w/ a logit link.

    - Example model formula: `glm(surv ~ size_1, data = my_surv_data, family = binomial())`
    

2. Growth (`g`): a linear model with a Normal error distribution.

    - Example model formula: `lm(size_2 ~ size_1, data = my_grow_data)`
    

3. Pr(flowering) (`f_r`): a generalized linear model w/ a logit link.

    - Example model formula: `glm(flower ~ size_1, data = my_repro_data, family = binomial())`
    

4. Seed production (`f_s`): a generalized linear model w/ log link.

    - Example model formula: `glm(seeds ~ size_1, data = my_flower_data, family = poisson())`
    

5. Recruit size distribution (`f_d`): a normal distribution w parameters `mu_fd` (mean) and `sd_fd` (standard deviation).

The example below assumes we've already fit our vital rate models from the raw data. In this example, the numbers are made up, but code that extracts the values you need from a real regression model is provided in comments.


```{r eval = FALSE}

# Load ipmr and get the parameter values. The data_list argument for define_kernel
# should hold every regression parameter and every constant used in the model.

library(ipmr)

data_list = list(s_int     = 2.2,   # coefficients(my_surv_mod)[1]
                 s_slope   = 0.25,  # coefficients(my_surv_mod)[2]
                 g_int     = 0.2,   # coefficients(my_grow_mod)[1]
                 g_slope   = 1.02,  # coefficients(my_grow_mod)[2]
                 sd_g      = 0.7,   # sd(resid(my_grow_mod))
                 f_r_int   = 0.003, # coefficients(my_pr_flower_mod)[1]
                 f_r_slope = 0.015, # coefficients(my_pr_flower_mod)[2]
                 f_s_int   = 1.3,   # coefficients(my_seed_mod)[1]
                 f_s_slope = 0.075, # coefficients(my_seed_mod)[2]
                 mu_fd     = 2,     # mean(recruit_data$size_next)
                 sd_fd     = 0.3)   # sd(recruit_data$size_next)

my_simple_ipm <- init_ipm('simple_di_det') %>%
  define_kernel(
    
    # Name of the kernel
    
    name      = "P_simple",
    
    # The type of transition it describes (e.g. continuous - continuous, discrete - continuous).
    # These must be specified for all kernels!
    
    family    = "CC",
    
    # The formula for the kernel. 
    
    formula   = s * g,
    
    # A named set of expressions for the vital rates it includes. 
    # note the use of user-specified functions here. Additionally, each 
    # state variable has a stateVariable_1 and stateVariable_2 internally defined
    # for the domain associated with it. Use these to distinguish between 
    # size/weight/etc at time t vs size/weight/etc at time t+1
    
    # Perform the inverse logit transformation to get survival probabilities
    # from your model. For examples on using predict(my_surv_mod,...),
    # see below.
    
    s         = 1 / (1 + exp(-(s_int + s_slope * dbh_1))), 
    
    # The growth model requires a function to compute the mean as a function of dbh.
    # The SD is a constant, so we don't need to define that in ... expression, 
    # just the data_list.
    
    g         = dnorm(dbh_2, mu_g, sd_g),
    mu_g      = g_int + g_slope * dbh_1,

    
    # Specify the constants in the model in the data_list. 
    
    data_list = data_list,
    states    = list(c('dbh')),
    
    # If you want to correct for eviction, set evict_cor = TRUE and specify an
    # evict_fun. ipmr provides truncated_distributions() to help.
    
    evict_cor = TRUE,
    evict_fun = truncated_distributions('norm',
                                        'g')
  ) %>%
  define_kernel('F_simple',
                formula   = f_r * f_s * f_d,
                family    = 'CC',
                
                # Inverse logit transformation for flowering probability
                # (because we used a logistic regression)
                
                f_r       = 1 / (1 + exp( - (f_r_int + f_r_slope * dbh_1))),
                
                # Exponential function for seed progression 
                # (because we used a Poisson)
                
                f_s       = exp(f_s_int + f_s_slope * dbh_1),
                
                # The recruit size distribution has no maternal effect for size,
                # so mu_fd and sd_fd are constants. These get passed in the 
                # data_list
                
                f_d       = dnorm(dbh_2, mu_fd, sd_fd),
                data_list = data_list,
                states    = list(c('dbh')),
                
                # Again, we'll correct for eviction in new recruits by
                # truncating the normal distribution.
                
                evict_cor = TRUE,
                evict_fun = truncated_distributions('norm',
                                                    'f_d')
  ) %>%
  
  # K kernels get their own special define_k function. Rather than use the formula
  # parameter, it simply takes named expressions for the format of the iteration
  # kernels. It can also take expressions showing how these kernels generate 
  # population states at t+1 as a function of population state at t - see examples
  # below for how to do that.
  
  define_k('K',
           K         = P_simple + F_simple,
           
           # This is a new family - at the moment all K's will get the
           # 'IPM' family
           
           family    = 'IPM',
           
           # This kernel has no additional parameters, so the data_list
           # is empty
           
           data_list = list(),
           states    = list(c('dbh')),
           # We've already corrected eviction in the sub-kernels, so there's no
           # need to do that here
           evict_cor = FALSE
  ) %>%
  # Next, we have to define the implementation details for the model. 
  # We need to tell ipmr how each kernel is integrated, what domain
  # it starts on (i.e. the size/weight/etc from above), and what domain
  # it ends on. In simple_* models, dom_start and dom_end will always be the same,
  # because we only have a single continuous state variable. General_*
  # models will be more complicated.
  
  define_impl(
    make_impl_args_list(
      kernel_names = c("K_simple", "P_simple", "F_simple"),
      int_rule     = rep("midpoint", 3),
      dom_start    = rep("dbh", 3),
      dom_end      = rep("dbh", 3)
    )
  ) %>%
  define_domains(
    dbh = c(0, # the first entry is the lower bound of the domain.
            50, # the second entry is the upper bound of the domain.
            100 # third entry is the number of meshpoints for the domain.
    ) 
  )  %>%
  make_ipm()


lambda_ipmr <- lambda(my_simple_ipm, comp_method = 'eigen')
w_ipmr      <- right_ev(my_simple_ipm)


```

If you're interested in seeing how `ipmr` output compares to models implemented by hand, there is an article on that [here](https://levisc8.github.io/ipmr/articles/sanity-checks.html). 

A more detailed introduction to `ipmr` is available in the Intro to ipmr article on this page. General IPM (ones with multiple continuous and/or discrete states) tutorials are also available there under the General IPMs article. That's all for now, happy modelling!
