---
output: 
  md_document:
    variant: markdown_github
---
[![Travis build status](https://travis-ci.org/levisc8/ipmr.svg?branch=master)](https://travis-ci.org/levisc8/ipmr)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/ipmr)](https://cran.r-project.org/package=ipmr)
[![Codecov test coverage](https://codecov.io/gh/levisc8/ipmr/branch/master/graph/badge.svg)](https://codecov.io/gh/levisc8/ipmr?branch=master)

# ipmr

`ipmr` is a package for implementing Integral Projection Models (IPMs) in _R_. It relies heavily on the mathematical syntax of the models, and does not try to abstract over the process of fitting vital rates. It is still very much a work in progress, and so models implemented using current code may not run in the not-so-distant future as more complicated methods are brought online and tweaks are made. Below is a brief overview of how `ipmr` classifies different model types followed by examples of how to implement those types in this framework. 

Note that this package **will not** help with the process of fitting vital rate models at all! That is a sufficiently different (and vast) question that we decided it was not within the scope of this project. This will only help you turn those regression models into an IPM without shooting yourself in the foot. Thus, everything that follows assumes you have parameterized those vital rate models and are now ready to begin implementing your model. 

Below is a brief overview of the package and some examples of how to implement models with it.

## Model classes

The first step of defining a model in `ipmr` (assuming all parameters have already been estimated) is to initialize the model using `init_ipm()`. This function takes a single argument: `model_class`. The `model_class` defines the basic infrastructure that will be available for subsequent analyses and helps make sure the kernels are correctly implemented from the underlying vital rates. `model_class` should be a character string with at least 3 (but possibly 4) entries separated by underscores (`_`). Below, the are the possible entries for each position.


- Position 1: `"simple"`/`"general"`

+ A.  **simple**: This describes an IPM with a single continuous state variable and no discrete stages. 

+ B.  **general**: This describes and IPM with either more than one continuous state variable, one or more discrete stages, or both of the above. Basically, anything other than an IPM with a single continuous state variable.

- Position 2: `"di"`/`"dd"`

+ A. **di**: This is used to denote a density-independent IPM.

+ B. **dd**: This is used to denote a density-dependent IPM.

- Position 3: `"det"`/`"stoch"`

+ A. **det**: This is used to denote a deterministic IPM. If this is used in the third position of `model_class`, there should not be a fourth entry.

+ B. **stoch**: This is used to denote a stochastic IPM. If this is used in the third position of `model_class`, there should always be a fourth entry. The two possibilities for the fourth are described next.

- Position 4: `"kern"`/`"param"`

+ A. **kern**: This describes an IPM with discretely varying parameters such that there values are known before the model is specified. This is usually the case with models that estimate random year/site effects and for which defining a multivariate joint distribution to sample parameters from is not desirable/needed. These models can be a bit more computationally efficient than the `param` alternative because all kernels can be constructed before the iteration procedure begins, as opposed to requiring reconstruction for every single iteration. 

+ B. **param**: This describes an IPM with parameters that are re-sampled from some distribution at each iteration of the model (usually a multivariate joint distribution). This can be a multivariate normal defined by covarying slopes and intercepts, or posterior distribution from a Bayesian model. All that is required is that the parameters for the distribution are specified and that the function that generates the parameters at each iteration returns named lists that correspond to the parameter names in the model. Jump down to the `"simple_di_stoch_param"` example for some inspiration in writing those.

With the type of model selected, the `model_class` becomes a string and the call to `init_ipm` is composed like so: `init_ipm(model_class = "position1_position_2_position3_position4")`. 

The following possibilities are currently or will become available in `ipmr` (bold text denotes development progress):

- Simple, density independent models: **Completed and ready**

1. `"simple_di_det"`

2. `"simple_di_stoch_kern"`

3. `"simple_di_stoch_param"`

- Simple, density dependent models: **Please be patient**

4. `"simple_dd_det"`

5. `"simple_dd_stoch_kern"`

6. `"simple_dd_stoch_param"`

- General, density independent models: **Completed and ready**

7. `"general_di_det"`

8. `"general_di_stoch_kern"`

9. `"general_di_stoch_param"`

-
General, density dependent models: **Please be patient**

10. `"general_dd_det"`

11. `"general_dd_stoch_kern"`

12. `"general_dd_stoch_param"`


## Examples for implemented IPM types

Simple density-independent deterministic, simple kernel-resampled stochastic, and simple parameter resampled stochastic models (`simple_di_det`, `simple_di_stoch_kern`, `simple_di_stoch_param`) are now functional. The `general_*` versions of these are also ready, and an introduction to them is available [here](https://levisc8.github.io/ipmr/articles/general-ipms.html). However, expect changes as more complicated methods are implemented! See below for an example of how to implement an IPM in this framework.

Next on the to-do list is to write generic functions for `lambda`, `right_ev` (right eigenvector), `left_ev` (left eigenvector), `sensitivity`, `elasticity`, and `plot`/`print` methods for all exported classes (`proto_ipm`, implemented `ipm` objects). After that, density dependent model classes will get implemented.

```{r eval = FALSE}

# Example of the setup for a simple IPM without density dependence or environmental
# stochasticity

library(ipmr)

data_list = list(s_int     = 2.2,
                 s_slope   = 0.25,
                 g_int     = 0.2,
                 g_slope   = 1.02,
                 sd_g      = 0.7,
                 f_r_int   = 0.003,
                 f_r_slope = 0.015,
                 f_s_int   = 1.3,
                 f_s_slope = 0.075,
                 mu_fd     = 2,
                 sd_fd     = 0.3)

s <- function(sv1, params) {
  1/(1 + exp(-(params[1] + params[2] * sv1)))
}


g <- function(sv1, sv2, params, L, U) {
  mu <- params[1] + params[2] * sv1
  ev <- (pnorm(U, mu, params[3]) - pnorm(L, mu, params[3]))
  dnorm(sv2, mean = mu, sd = params[3]) / ev
}

f_r <- function(sv1, params) {
  1/(1 + exp(-(params[1] + params[2] * sv1)))
}

f_s <- function(sv1, params) {
  exp(params[1] + params[2] * sv1)
}

f_d <- function(sv2, params) {
  dnorm(sv2, mean = params[1], sd = params[2])
}

fec <- function(sv1, sv2, params, L, U) {
  ev <- (pnorm(U, params[5], params[6]) - pnorm(L, params[5], params[6]))
  f_r(sv1, params[1:2]) * f_s(sv1, params[3:4]) * (f_d(sv2, params[5:6] / ev))
}

b <- seq(0, 50, length.out = 101)
d1 <- (b[2:101] + b[1:100]) * 0.5
h <- d1[3] - d1[2]

domains <- expand.grid(list(d2 = d1, d1 = d1))

G <- g(domains$d2,
       domains$d1,
       params = c(data_list$g_int,
                  data_list$g_slope,
                  data_list$sd_g),
       L = 0,
       U = 50)



S <- s(d1, c(data_list$s_int, data_list$s_slope))

P <- h * t(S * t(G))

Fm <- h * fec(domains$d2,
              domains$d1,
              params = unlist(data_list[6:11]),
              L = 0, U = 50)

K <- P + Fm

K <- matrix(K, nrow = 100, ncol = 100, byrow = TRUE)


lambda_usr <- Re(eigen(K)$values[1])
w <- Re(eigen(K)$vectors[ , 1])


# User specified functions can be passed to make_ipm(usr_funs = list(my_fun = my_fun)).
# inv_logit is a simple example, but more complicated ones can be specified as well. 

inv_logit <- function(int, slope, sv) {
  return(1/(1 + exp(-(int + slope * sv))))
}


impl_args <- make_impl_args_list(
  kernel_names = c("K", "P", "F"),
  int_rule     = rep("midpoint", 3),
  dom_start    = rep("dbh", 3),
  dom_end      = rep("dbh", 3)
)
states <- list(c('dbh'))

x <- init_ipm('simple_di_det') %>%
  define_kernel(
    
    # Name of the kernel
    
    name      = "P",
    
    # The type of transition it describes (e.g. continuous - continuous, discrete - continuous)
    
    family    = "CC",
    
    # The formula for the kernel. s_g_mult() is a helper function to make sure the
    # survival vector is correctly aligned with the kernel.
    
    formula   = s_g_mult(s, g),
    
    # A named set of expressions for the vital rates it includes. 
    # note the use of user-specified functions here. Additionally, each 
    # state variable has an L_StateVariable and U_StateVariable_ internally defined
    # for the domain associated with it, so you can reference those when 
    # when specifying your own functions. 
    
    s         = inv_logit(s_int, s_slope, dbh_1), 
    g         = dnorm(dbh_2, mu_g, sd_g),
    mu_g      = g_int + g_slope * dbh_1,

    data_list = data_list,
    states    = states,
    evict     = TRUE,
    evict_fun = truncated_distributions('norm',
                                        'g')
  ) %>%
  define_kernel('F',
                formula   = f_r * f_s * f_d,
                family    = 'CC',
                f_r       = inv_logit(f_r_int, f_r_slope, dbh_1),
                f_s       = exp(f_s_int + f_s_slope * dbh_1),
                f_d       = dnorm(dbh_2, mu_fd, sd_fd),
                data_list = data_list,
                states    = states,
                evict     = TRUE,
                evict_fun = truncated_distributions('norm',
                                                    'f_d')
  ) %>%
  
  # K kernels get their own special define_k function. Rather than use the formula
  # parameter, it simply takes ..., allowing you to specify the form of the iteration
  # kernel and the population vector at T + 1 simulataneously. This is likely to
  # be more useful for stochastic simulation models than deterministic models
  
  define_k('K',
           K         = P + F,
           family    = 'IPM',
           data_list = list(),
           states    = states,
           evict     = FALSE) %>%
  define_impl(impl_args) %>%
  define_domains(
    dbh = c(0, # the first entry is the lower bound of the domain.
            50, # the second entry is the upper bound of the domain.
            100 # third entry is the number of meshpoints for the domain.
    ) 
  )  %>%
  make_ipm(usr_funs = list(inv_logit = inv_logit))


lambda_ipmr <- Re(eigen(x$iterators$K)$values[1])
w_ipmr      <- Re(eigen(x$iterators$K)$vectors[ , 1])


lambda_ipmr - lambda_usr


```

## Simple, density independent, stochastic kernel resampled models

These models are typically the result of vital rate models that are fit in a mixed effects framework (e.g. multiple sites or multiple years of data). They have a special syntax that mirrors the mathematical notation of these models (and has the side effect of saving you a considerable amount of copying/pasting/typing in general).

The syntax uses a `name_hierarchicalVariable` notation. These names are automatically expanded to include the multiple levels of the hierarchical variable. For example, the P kernel for a model with 5 years of data could be `name`'d `P_yr`, and the underlying vital rates would be denoted `vitalRate_yr`. The parameter values in the `data_list` are the exception to this - they must be labeled with the actual values that the hierarchical variable takes.

The example below simulates a lifecycle where reproduction is always fatal. Thus, the survival term also includes the probability of reproducing. This is meant to (hopefully) demonstrate the flexibility of the framework.

```{r eval = FALSE}

# rlang is a useful shortcut for splicing named values into lists. purrr is used to manipulate said lists.
# This is intended to simulate a monocarpic perennial life history where flowering is always fatal.
# Note that this means the survival function also includes the probability of reproduction function.

library(ipmr)
library(purrr)

# define functions for target ipm


# Survival - logistic regression
s <- function(sv1, params, r_effect) {
  1/(1 + exp(-(params[1] + params[2] * sv1 + r_effect))) *
    (1 - f_r(sv1, params[3:4]))
}

# Growth
g <- function(sv1, sv2, params, r_effect, L, U) {
  mu <- params[1] + params[2] * sv1 + r_effect
  ev <- pnorm(U, mu, params[3]) - pnorm(L, mu, params[3])
  dnorm(sv2, mean = mu, sd = params[3]) / ev
}

# probability of reproducing
f_r <- function(sv1, params) {
  1/(1 + exp(-(params[1] + params[2] * sv1)))
}

# offspring production
f_s <- function(sv1, params, r_effect) {
  exp(params[1] + params[2] * sv1 + r_effect)
}

# offspring size distribution
f_d <- function(sv2, params, L, U) {
  ev <- pnorm(U, params[1], params[2]) - pnorm(L, params[1], params[2])
  dnorm(sv2, mean = params[1], sd = params[2]) / ev
}

# constructor function for the F kernel
fec <- function(sv1, sv2, params, r_effect, L, U) {
  f_r(sv1, params[1:2]) * f_s(sv1, params[3:4], r_effect) * f_d(sv2, params[5:6], L, U)
}


set.seed(50127)


# Define some fixed parameters
data_list = list(
  s_int     = 1.03,
  s_slope   = 2.2,
  g_int     = 8,
  g_slope   = 0.92,
  sd_g      = 0.9,
  f_r_int   = 0.09,
  f_r_slope = 0.05,
  f_s_int   = 0.1,
  f_s_slope = 0.005,
  mu_fd     = 9,
  sd_fd     = 2
)

# Now, simulate some random intercepts for growth, survival, and offspring production

g_r_int   <- rnorm(5, 0, 0.3)
s_r_int   <- rnorm(5, 0, 0.7)
f_s_r_int <- rnorm(5, 0, 0.2)

nms <- paste("r_", 1:5, sep = "")

names(g_r_int)   <- paste('g_', nms, sep = "")
names(s_r_int)   <- paste('s_', nms, sep = "")
names(f_s_r_int) <- paste('f_s_', nms, sep = "")

# Each set of parameters is converted to a named list. The names should match
# the variables referenced in each define_kernel()/define_k() call.

g_params   <- as.list(g_r_int)
s_params   <- as.list(s_r_int)
f_s_params <- as.list(f_s_r_int)

# purrr::splice combines each separate list into one without creating any 
# additional depth to it. This keeps things a bit tidier.

params     <- splice(data_list, g_params, s_params, f_s_params)


b   <- seq(0.2, 40, length.out = 101)

sv1 <- (b[2:101] + b[1:100]) * 0.5

domains <- expand.grid(list(d2 = sv1, d1 = sv1))

h   <- sv1[2] - sv1[1]

# repetitive to demonstrate the typical kernel construction process.

g_1 <- g(domains$d2, domains$d1,
         params = c(params$g_int,
                    params$g_slope,
                    params$sd_g),
         r_effect = params$g_r_1,
         L = 0.2,
         U = 40)

g_2 <- g(domains$d2, domains$d1,
         params = c(params$g_int,
                    params$g_slope,
                    params$sd_g),
         r_effect = params$g_r_2,
         L = 0.2,
         U = 40)

g_3 <- g(domains$d2, domains$d1,
         params = c(params$g_int,
                    params$g_slope,
                    params$sd_g),
         r_effect = params$g_r_3,
         L = 0.2,
         U = 40)


g_4 <- g(domains$d2, domains$d1,
         params = c(params$g_int,
                    params$g_slope,
                    params$sd_g),
         r_effect = params$g_r_4,
         L = 0.2,
         U = 40)

g_5 <- g(domains$d2, domains$d1,
         params = c(params$g_int,
                    params$g_slope,
                    params$sd_g),
         r_effect = params$g_r_5,
         L = 0.2,
         U = 40)

s_1 <- s(sv1, c(params$s_int, params$s_slope,
                params$f_r_int, params$f_r_slope), params$s_r_1)
s_2 <- s(sv1, c(params$s_int, params$s_slope,
                params$f_r_int, params$f_r_slope), params$s_r_2)
s_3 <- s(sv1, c(params$s_int, params$s_slope,
                params$f_r_int, params$f_r_slope), params$s_r_3)
s_4 <- s(sv1, c(params$s_int, params$s_slope,
                params$f_r_int, params$f_r_slope), params$s_r_4)
s_5 <- s(sv1, c(params$s_int, params$s_slope,
                params$f_r_int, params$f_r_slope), params$s_r_5)

P_1 <- t(s_1 * t(g_1)) * h
P_2 <- t(s_2 * t(g_2)) * h
P_3 <- t(s_3 * t(g_3)) * h
P_4 <- t(s_4 * t(g_4)) * h
P_5 <- t(s_5 * t(g_5)) * h

# These are not corrected for eviction, but they probably should be

F_1 <- h * fec(domains$d2, domains$d1,
               params = unlist(params[6:11]),
               r_effect = params$f_s_r_1,
               L = 0.2,
               U = 40)

F_2 <- h * fec(domains$d2, domains$d1,
               params = unlist(params[6:11]),
               r_effect = params$f_s_r_2,
               L = 0.2,
               U = 40)

F_3 <- h * fec(domains$d2, domains$d1,
               params = unlist(params[6:11]),
               r_effect = params$f_s_r_3,
               L = 0.2,
               U = 40)
F_4 <- h * fec(domains$d2, domains$d1,
               params = unlist(params[6:11]),
               r_effect = params$f_s_r_4,
               L = 0.2,
               U = 40)
F_5 <- h * fec(domains$d2, domains$d1,
               params = unlist(params[6:11]),
               r_effect = params$f_s_r_5,
               L = 0.2,
               U = 40)

K_1 <- (P_1 + F_1) %>%
  matrix(nrow = 100, ncol = 100, byrow = TRUE)
K_2 <- (P_2 + F_2) %>%
  matrix(nrow = 100, ncol = 100, byrow = TRUE)
K_3 <- (P_3 + F_3) %>%
  matrix(nrow = 100, ncol = 100, byrow = TRUE)
K_4 <- (P_4 + F_4) %>%
  matrix(nrow = 100, ncol = 100, byrow = TRUE)
K_5 <- (P_5 + F_5) %>%
  matrix(nrow = 100, ncol = 100, byrow = TRUE)

sys <- list(K_1 = K_1,
            K_2 = K_2,
            K_3 = K_3,
            K_4 = K_4,
            K_5 = K_5)

eigen_sys <- lapply(sys, eigen)

lambdas <- vapply(eigen_sys, function(x) Re(x$values[1]), numeric(1))
ws      <- vapply(eigen_sys, function(x) Re(x$vectors[ , 1]), numeric(100))



## ipmr version

# define the levels of the hierarchical variable and save them in a named
# list that corresponds to the suffix in the kernel notation

hier_levels <- list(yr = 1:5)

# additional usr_funs to be passed into make_ipm()

inv_logit <- function(sv, int, slope) {
  return(
    1/(1 + exp(-(int + slope * sv)))
  )
}

inv_logit_r <- function(sv, int, slope, r_eff) {
  return(
    1/(1 + exp(-(int + slope * sv + r_eff)))
  )
}

pois_r <- function(sv, int, slope, r_eff) {
  return(
    exp(
      int + slope * sv + r_eff
    )
  )
}

# The "model_class" argument is now changed to reflect a different model type

monocarp_sys <- init_ipm('simple_di_stoch_kern') %>%
  define_kernel(
    
    # The yr suffix is appended to the kernel name and the parameter names
    # within each vital rate expression. ipmr substitutes in the hier_levels
    # for each suffix occurrence, thus changing P_yr in P_1, P_2, P_3, P_4, P_5,
    # s_yr in s_1, s_2, s_3, s_4, and s_5. s_r_yr is converted to s_r_1, s_r_2, 
    # etc. In the case of s_r_yr, provided that the names in the data_list match
    # the expanded names, all will go well!
    
    name             = 'P_yr',
    formula          = s_g_mult(s_yr, g_yr) ,
    family           = "CC",
    s_yr             = inv_logit_r(ht_1, s_int, s_slope, s_r_yr) * 
      (1 - inv_logit(ht_1, f_r_int, f_r_slope)),
    g_yr             = dnorm(ht_2, mu_g_yr, sd_g),
    mu_g_yr          = g_int + g_slope * ht_1 + g_r_yr,
    data_list        = params,
    states           = list(c('ht')),
    has_hier_effs    = TRUE,
    levels_hier_effs = hier_levels,
    evict            = TRUE,
    evict_fun        = truncated_distributions("norm", "g_yr")
  ) %>%
  define_kernel(
    name             = "F_yr",
    formula          = f_r * f_s_yr * f_d,
    family           = "CC",
    f_r              = inv_logit(ht_1, f_r_int, f_r_slope),
    f_s_yr           = pois_r(ht_1, f_s_int, f_s_slope, f_s_r_yr),
    f_d              = dnorm(ht_2, mu_fd, sd_fd),
    data_list        = params,
    states           = list(c('ht')),
    has_hier_effs    = TRUE,
    levels_hier_effs = hier_levels,
    evict            = TRUE,
    evict_fun        = truncated_distributions("norm", "f_d")
  ) %>%
  define_k(
    name             = 'K_yr',
    K_yr             = P_yr + F_yr,
    family           = "IPM",
    data_list        = params,
    states           = list(c("ht")),
    has_hier_effs    = TRUE,
    levels_hier_effs = hier_levels
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c("K_yr", "P_yr", "F_yr"),
      int_rule     = rep("midpoint", 3),
      dom_start    = rep("ht", 3),
      dom_end      = rep("ht", 3)
    )
  ) %>%
  define_domains(ht = c(0.2, 40, 100)) %>%
  make_ipm(usr_funs = list(inv_logit   = inv_logit,
                           inv_logit_r = inv_logit_r,
                           pois_r      = pois_r))



lambda_ipmr <- vapply(monocarp_sys$iterators, 
                      function(x) Re(eigen(x)$values[1]), 
                      numeric(1))

lambda_ipmr - lambdas


```

## Simple, density independet, parameter resampled models

These models are stochastic, but rather than building the iteration kernels first and then iterating them - distributions for varying parameters are specified ahead of time and then resampled at each iteration. `ipmr` is careful to only evaluate each expression once per iteration, allowing you to specify parameters sampled from a joint distribution with a single expression. Their notation *usually* does not require suffix notation, though this example uses it to make clear which parameters are varying and which are fixed. All that's required is that: 

1. the function returns a named list of parameter values and 

2. the names of the list values and the names used in the vital rate/kernel expressions match. 

The user-specified function should go into `define_env_state()` call. The values can then be referenced using `list_name$parameter_name` notation in the vital rate expressions.

This example uses the `rmvnorm` function from the `mvtnorm` package to demonstrate how one might sample from a joint distribution of parameters. One could also pass a function that samples from a posterior distribution of a Bayesian model, and pass the associated posterior to the `data_list` argument in `define_env_state` to propagate parameter uncertainty throughout the model.

```{r eval = FALSE}

library(ipmr)
library(mvtnorm)

# Define the fixed parameters in a list

set.seed(2123008)

data_list <- list(s_slope   = 0.3,
                  g_slope   = 0.99,
                  g_sd      = 0.2,
                  f_r_slope = 0.03,
                  f_s_slope = 0.001,
                  f_d_mu    = 1.1,
                  f_d_sd    = 0.1)

# Simulate some random parameters to define the multivariate distribution
# to sample from. In user-specified models, these will likely come from mixed
# effects models fit to real data.
r_means <- c(s_int_yr   = 0.8,
             g_int_yr   = 0.1,
             f_r_int_yr = 0.3,
             f_s_int_yr = 0.4)

r_sigma <- runif(16)
r_sigma <- matrix(r_sigma, nrow = 4)

r_sigma <- r_sigma %*% t(r_sigma) # make symmetrical - most users should skip this step.

iterations <- 10

# Specify an initial population vector. This can also be done in the call to
# define_pop_state

init_pop_vec <- runif(100)

# Again, we can define our own functions and pass them into calls to make_ipm
inv_logit <- function(int, slope, sv1) {
  1/(1 + exp(-(int + slope * sv1)))
}

# In this case, we define a wrapper function around rmvnorm to make sure
# that the names of the returned values match those in the vital rate
# expressions.

mvt_wrapper <- function(r_means, r_sigma, nms) {
  out <- mvtnorm::rmvnorm(1, r_means, r_sigma) %>%
    as.list()
  
  names(out) <- nms
  return(out)
}

param_resamp_model <- init_ipm('simple_di_stoch_param') %>%
  define_kernel(
    name    = 'P',
    formula = s_g_mult(s, g),
    family  = 'CC',
    
    # Note here that the growth and survival intercepts are prefaced with env_params$...
    # This is because the mvt_wrapper function is named env_params in the call
    # to define_env_state(). env_params is not a reserved word, so you can call
    # it whatever you like.
    
    g_mu          = env_params$g_int_yr + g_slope * surf_area_1,
    s             = inv_logit(env_params$s_int_yr, s_slope, surf_area_1),
    g             = dnorm(surf_area_2, g_mu, g_sd),
    data_list     = data_list,
    states        = list(c('surf_area')),
    has_hier_effs = FALSE,
    evict         = TRUE,
    evict_fun     = truncated_distributions("norm", "g")
  ) %>%
  define_kernel(
    name          = 'F',
    formula       = f_r * f_s * f_d,
    family        = 'CC',
    
    # Note the insertion of env_params$param_name here as well
    
    f_r           = inv_logit(env_params$f_r_int_yr, f_r_slope, surf_area_1),
    f_s           = exp(env_params$f_s_int_yr + f_s_slope * surf_area_1),
    f_d           = dnorm(surf_area_2, f_d_mu, f_d_sd),
    
    data_list     = data_list,
    states        = list(c('surf_area')),
    has_hier_effs = FALSE,
    evict         = TRUE,
    evict_fun     = truncated_distributions("norm", "f_d")
  ) %>%
  define_k(
    name = 'K',
    
    # Note that here, we specify both the form of the iteration kernel and
    # the iteration procedure. right_mult() is a helper function to make sure
    # population states are multiplied by the kernels correctly.
    
    K               = P + F,
    n_surf_area_t_1 = right_mult(K, n_surf_area_t),
    
    family          = 'IPM',
    data_list       = data_list,
    states          = list(c('surf_area')),
    has_hier_effs   = FALSE,
    evict           = FALSE
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c('P', "F", "K"),
      int_rule     = rep('midpoint', 3),
      dom_start    = rep('surf_area',3),
      dom_end      = rep('surf_area', 3)
    )
  ) %>%
  define_domains(surf_area = c(0, 10, 100)) %>%
  define_env_state(
    env_params = mvt_wrapper(r_means, r_sigma, nms = c('s_int_yr',
                                                       'g_int_yr',
                                                       'f_r_int_yr',
                                                       'f_s_int_yr')),
    data_list = list(
      r_means = r_means,
      r_sigma = r_sigma
    )
  ) %>%
  define_pop_state(
    pop_vectors = list(n_surf_area_t = init_pop_vec),
  ) %>%
  make_ipm(usr_funs   = list(inv_logit = inv_logit,
                             mvt_wrapper = mvt_wrapper),
           iterate    = TRUE,
           iterations = 10)

pop_state_ipmr <- param_resamp_model$pop_state$pop_state_surf_area
lambda_ipmr    <- numeric(iterations)

for(i in seq(2, dim(pop_state_ipmr)[2], 1)) {
  
  lambda_ipmr[(i - 1)] <- sum(pop_state_ipmr[ ,i]) / sum(pop_state_ipmr[ ,(i - 1)])
  
}

lambda_ipmr

```
