---
title: "ipmr: Flexibly implement Integral Projection Models in R"
output:  
  word_document:
    toc: no
fig_caption: yes
---

Sam C. Levin $^{\ast1,2,3}$, Aldo Compagnoni $^{1,2,3}$, Roberto Salguero-Gomez $^3$, Dylan Z. Childs $^4$, Sanne Evers $^{1,2}$, Tiffany M. Knight $^{1,2,5}$

$^1$Institute of Biology, Martin Luther University Halle-Wittenberg, Am Kirchtor 1, 06108 Halle (Saale), Germany

$^2$German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Deutscher Platz 5e, 04103 Leipzig, Germany

$^3$Oxford Zoology

$^4$Sheffield

$^5$UFZ



$^*$Corresponding Author:

Sam C Levin

Deutscherplatz 5e, 04103 Leipzig, Germany

email: <levisc8@gmail.com>



Target journal: Methods in Ecology and Evolution, submission type: Application. Word limit: 3000



[Formatting guidelines](https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/author-guidelines) (in case you care)

Keywords: integral projection model, population dynamics, structured populations, life history

# **Abstract**

1. Integral projection models (IPMs) have become an important tool for studying the demography of populations structured by one or more continuous state variables. They are used to investigate questions ranging from single species conservation and management strategies for invasive species to understanding selective pressures in Soay sheep. _R_ is a commonly used data analysis tool for ecologists. However, to date, there has only been one package for _R_ that implements integral projection models - `IPMpack`. We offer an alternative, more flexible implementation for building IPMs that can handle a wider variety of applications.

2. `ipmr` can implement determinstic or stochastic, density independent IPMs for discretely or continuously varying environments. The package operates at a much lower level of abstraction than previous efforts. Our framework relies on the mathematical notation of the models, and completely decouples the model parameterization step from the model implementation step. The latter point substantially increases `ipmr`'s flexibility to model complex life cycles. 

3. The `ipmr` package can handle a wider variety of models than currently existing software, including discretely and continuously stochastic environments, and multiple continuous and or discrete state variables. Specific methods for models with individuals cross-classified by age and size are also available. Furthermore, the package provides methods for basic analyses and visualization, and tools to facilitate more complex analyses.

4. `ipmr` is far more flexible than existing packages for implementing integral projection models. The common underlying data structure used to generate models also presents opportunities to enhance reproducibility in the field. Furthermore, the package could be a template that allows researchers to represent IPMs in concise, text format databases that rely on functional forms and parameters values, rather than raw data, to represent models. 

```{r echo = FALSE, message = FALSE, warning=FALSE}

library(dplyr)
library(purrr)
library(ggplot2)

ipm_pubs <- read.csv('../../data-raw/padr-pubs-feb-2020.csv',
                     stringsAsFactors = FALSE)

ipm_id <- pmap_chr(.l = data.frame(a = ipm_pubs$Authors,
                                   b = ipm_pubs$Journal, 
                                   c = ipm_pubs$Year),
               .f = function(a, b, c) paste(a, b, c, sep = "_")) 

ipm_pubs <- cbind(ipm_pubs, ipm_id)
ipm_pubs$Year <- as.integer(ipm_pubs$Year)

pub_tot  <- length(unique(ipm_id))

spec_tot <- length(unique(ipm_pubs$Species))

cdb_fetch <- function(cdb) {
  # get url or path
  if (tolower(cdb) == 'comadre') {
    path <- url('https://compadre-db.org/Data/ComadreDownload')
  } else if (tolower(cdb) == 'compadre') {
    path <- url('https://compadre-db.org/Data/CompadreDownload')
  } else {
    path <- path.expand(cdb)
  }
  
  # fetch and load
  env <- new.env()
  x <- load(path, env)[1]
  dbFetch <- env[[x]]

  # Deal with differences between s4 and s3 versions of database

  if(inherits(dbFetch, 'list')) {
    dbFetch <- dbFetch[[1]]
  } else if(inherits(dbFetch, "CompadreDB")) {
    dbFetch <- dbFetch@data
  } else {
    stop("Cannot recognize class of currently fetched com(p)adre object",
         call. = FALSE)
  }


  return(dbFetch)

}

cpd <- cdb_fetch('compadre')

mpm_id<- pmap_chr(.l = data.frame(a = cpd$Authors,
                                  b = cpd$Journal, 
                                  c = cpd$YearPublication),
               .f = function(a, b, c) paste(a, b, c, sep = "_")) 

cpd <- cbind(cpd, mpm_id)

cpd$YearPublication <- as.integer(cpd$YearPublication)

cpd_tot_pubs <- cpd %>% 
  filter(!duplicated(mpm_id)) %>% 
  group_by(YearPublication) %>%
  summarise(n_tot = n()) %>%
  ungroup() %>%
  arrange(YearPublication) %>%
  filter(!is.na(YearPublication))

cpd_tot_pubs <- mutate(cpd_tot_pubs, run_sum = cumsum(n_tot),
                       Database = "Compadre_MPM_Database") %>%
  setNames(c(
    "Year",
    "Number_per_Year",
    "Cumulative_Publications",
    "Database"
  ))

pdr_tot_pubs <- ipm_pubs %>%
  filter(!duplicated(ipm_id)) %>%
  group_by(Year) %>%
  summarise(n_tot = n()) %>%
  ungroup() %>%
  arrange(Year)
  
pdr_tot_pubs <- mutate(pdr_tot_pubs, run_sum = cumsum(n_tot),
                       Database = "Padrino_IPM_Database") %>%
  setNames(c(
    "Year",
    "Number_per_Year",
    "Cumulative_Publications",
    "Database"
  ))  %>%
  filter(!is.na(Year))

all_pubs <- rbind(pdr_tot_pubs, cpd_tot_pubs)
all_pubs$Year <- as.integer(all_pubs$Year)

```


# **Introduction**

Integral projection models (IPMs) are an important tool for ecologists studying population dynamics in discrete time. Since their introduction in Easterling, Ellner & Dixon's (2000) paper, there have been over `r pub_tot` publications using IPMs to address a variety of questions ranging from invasive species population dynamics (e.g. Crandall & Knight 2017) to evolutionary stable strategies to endangered species conservation (Childs et al. 2011, Coulson 2012). In the plant kingdom alone, there are currently at least `r spec_tot` plant species for which an IPM exists (Levin et al. unpublished). Advantages of using an IPM include (i) they allow for continuously distributed state variables to describe vital rates and population structure, (ii) the ability to mix discrete and continuous states, and (iii) allow for efficient parameter estimation with widely used regression methods. Since the first IPM publication, methods to environmental stochasticity, spatial dynamics, and density dependence have been developed (Ellner & Rees 2006, Adler et al. 2010, Ellner et al. 2016).

IPMs are flexible models that often, but not exclusively, use individual level longitudinal data to determine the functional forms of state-fate relationships such as survival, growth/maturation, and (a)sexual reproduction. They are built using regression models of vital rates on a state variable that is related to individual fitness (usually a measure of size, e.g. diameter at breast height for trees, body mass in animals). Regression modelling is a standard tool for ecologists, and the field is mature enough to accomodate many different types of data (e.g. Wood 2011, Bates et al. 2015). Parameter estimation for regression models is usually more efficient than estimating individual matrix transition parameters (Ellner & Rees 2006). There are usually fewer parameters to estimate with the same amount of data (in theory, one could estimate matrix transitions with regression models, but this is not often done in practice). This flexibility has lead to a rapid increase in their appearance in the literature (Figure 1). This rise in popularity also spurred the creation of software tools and guides that assisted in their implementation and analysis (Metcalf et al. 2013, Merow et al. 2014, Ellner et al. 2016). These range from _R_ scripts with detailed annotations to a complete R package.

```{r figure 1, echo = FALSE, fig.height = 6, fig.width = 8, message = FALSE, warning = FALSE, fig.cap = "Figure 1: Cumulative number of publications that make use of Matrix Projection Models (MPMs) and Integral Projection Models (IPMs) (A) and number of publications per year for each type of model (B). IPMs have been adopted rapidly since the first publication describing them in 2000. Unfortunately, software packages to assist with their implementation have not kept pace with theoretical advancements."}
 
library(gridExtra)
library(grid)

run_sum_plot <- ggplot(all_pubs,
       aes(x = Year,
           y = Cumulative_Publications)) +
  geom_line(aes(color = Database),
            size = 1.25) +
  theme_bw() +
  scale_color_manual(breaks = c("Compadre_MPM_Database",
                                "Padrino_IPM_Database"),
                     labels = c("Compadre MPM Database",
                                "Padrino IPM Database"),
                     values = viridis::inferno(2,
                                               begin = 0,
                                               end = 0.5,
                                               direction = -1)) +
  annotate("text", x = 1965, y = 600, label = "A", size = 10) +
  theme(legend.position = "none") +
  ylab("Cumulative Publications")

ann_pubs <- ggplot(all_pubs,
                   aes(x = Year)) +
  geom_col(aes(y = Number_per_Year,
               fill = Database)) +
  theme_bw() +
  scale_fill_manual(breaks = c("Compadre_MPM_Database",
                               "Padrino_IPM_Database"),
                    labels = c("Compadre MPM Database",
                               "Padrino IPM Database"),
                    values = viridis::inferno(2,
                                               begin = 0,
                                               end = 0.5,
                                               direction = -1)) +
  annotate("text", x = 1965, y = 50, label = "B", size = 10) +
  theme(legend.position = "bottom") +
  ylab("Number per Year")

grid.arrange(run_sum_plot, ann_pubs)

```


The only existing _R_ package that implements IPMs is `IPMpack`. `IPMpack` allows users to go from raw data to a set of projection kernels with just a few lines of code. However, this package is no longer actively developed, and so hasn't kept pace with more recent developments in IPM theory. `ipmr` fills this gap in a number of ways.  First, the package is explictly designed to work with more types of IPMs, and can handle models with individuals classified by a mixture of any number of continuously and discretely distributed state variables. Furthermore, it introduces specific classes and methods to deal with both discretely and continuously varying stochastic environments. Finally, the decoupling of the parameterization (i.e. regression model fitting) and implementation steps (i.e. kernel construction) of model building allows users to specify models of any functional form they desire.

First, we review the form of an IPM and some associated terms. Next, we delve into two case studies that illustrate potential applications of `ipmr`. The first case study introduces the `ipmr` syntax with a simple IPM. Note that the simple model could also be implemented in `IPMpack`. The second is intended to demonstrate the ways in which `ipmr`'s capabilities extend beyond those in `IPMpack` using an age- and size-structured model. The Online Supplementary Materials contains code for both, as well as examples of how to derive a couple life history traits once the model is implemented.  We then discuss additional opportunities this framework provides.

# Defining an IPM and terminology 

Before describing `ipmr` in greater detail, we first need to define the various components of the model and some terminology. An IPM is comprised of a projection kernel (denoted $K$), and a trait distribution function. The latter describes the frequency of a trait value in a population, and is analogous to the population vector in a matrix projection model. The model can be written as follows:

1. $n(z', t+1) = \int_L^UK(z',z)n(z,t)\mathrm{dz}$

Here, $n(z, t)$ and $n(z',t+1)$ are the distribution functions for some trait, $z$ at time $t$ and $t+1$. $L,U$ are the lower and upper bounds of values that a trait can have, and $K(z',z)$ is the projection kernel. The integral over $n(z,t)$ is the total population size. In the interest of making the model more biologically interpretable, the $K(z',z)$ kernel is usually decomposed into sub-kernels. For example, a model that captures survival/maturation transitions and sexual reproduction may be written as follows:

2. $K(z',z) = P(z',z) + F(z',z)$

Here, $P(z',z)$ describes transitions due to survival and growth/maturation of existing individuals, and $F(z',z)$ describes transitions that create new individuals through sexual reproduction. The sub-kernels themselves are typically generated with functions parameterized by regression models that relate the individuals trait value $z$ at time $t$ to some demographic outcome at $t+1$. For example, the $P$ kernel for a plant may contain two regression models: (i) a logistic regression of survival on plant size, and (ii) a regression of size at $t+1$ on size at $t$. We refer to these different scales (regressions and parameters, sub-kernels, and projection kernels) as levels of the model hierarchy. There have been numerous extensions of IPM theory. Examples include outlining a general, deterministic IPM that allows for a mixture of multiple discrete and/or continuous state variables (Ellner & Rees 2006), stochastically varying parameters to generate unique kernels (Rees & Ellner 2009), and spatial dynamics (Jongejans et al. 2011).

Analytical solutions to the integrals in Equation 1 are usually too complicated (or not possible). Numerical approximations of these integrals are typically quite easy to compute though. Examples of integration rules are the midpoint rule and the trapezoid rule for integration, though more complicated and precise methods have also been used (Zuidema et al. 2010). A matrix multiplication of the discretized projection kernel and the discretized trait distribution then generates a new trait distribution, a process referred to as model iteration. 

A critical aspect of `ipmr`'s functionality is the distinction between simple and general IPMs. A simple model is any IPM that makes use of one, and only one, continuous state variable, and models no additional structure in the population (i.e. contains no discrete states). Equations 1 and 2 represent a simple IPM. A general IPM models one or more continuous state variables, and/or discrete states. Examples of discrete state variables are number of seeds in a seed bank, or number of insect eggs laid that are not yet hatched. A more comprehensive defintion is given by Ellner & Rees (2006) and Ellner, Rees & Childs (2016). 

# Design and usage

The `ipmr` model construction syntax is designed to mimic the mathematical notation of the IPM (Tables 1 & 2). The package is powered by the `rlang` implementation of quasi-quotation (Henry & Wickham 2019). On the user side, the package works by building up expressions that reference each other at higher and higher levels in the model hierarchy (i.e. regression models, sub-kernels, projection kernels). `ipmr` is relatively dependency-free, requiring only `rlang`, `purrr` (Henry & Wickham 2020), and `magrittr` (Bache & Wickham 2014). The model definition functions (`define_kernel`, `define_k`) can handle any valid R code in the expressions that are passed to them, so user-specified functions, in addition to ones included in other packages can all be used in vital rate and kernel expressions (e.g. `predict.*` methods for complicated vital rate models). Additionally, there are helper functions that define the kernel implementation, the initial state of the system, and optionally, how the system changes over the course of a simulation (e.g. environmental fluctuations and/or correlations). Finally, the package implements a suffix-based syntax for symbolically representing models with one or more grouping variables (e.g. data from multiple sites, years, or an age-structured model, see Table 2 and code in the ESM for details). All functions, methods, and classes are documented extensively on the package's website (https://levisc8.github.io/ipmr/), and in the package's vignettes and help files. Figure 2 shows a generic IPM workflow from collecting the raw data to biological inference, highlighting the stages at which `ipmr` is useful. 

```{r figure 2, fig.cap = "The first step of any IPM is collecting individual level, longitudinal data in the field (1). After collecting and cleaning the data, state-fate relationships need to be modelled. Additionally, other constant parameters may be defined, such as seed germination probabilities (2). In order to work with `ipmr`'s functions, the next step is to generate a named list of parameter values, such that each name in the list corresponds to a variable in the model (3). The research question will determine whether a deterministic or stochastic, general or simple model is necessary, and this is specified with *init_ipm()*. Models with individuals classified by age and size can also be specified here (4). Next, expressions that represent each vital rate function are defined, and these are combined into sub-kernels using *define_kernel()*. Simpler models can be generated faster using the mathematical form of the mean (and possibly variance) functions from the regression models. However, more complicated functional and/or semi- or non-parametric functional forms can be written using predict() methods from any package the user desires (5). After defining the kernel, the next step is to decide on integration rules, domains, and optionally, whether to define an initial population state and environmental state (6). This concludes the model definition stage, and generates a *proto_ipm*. This is then passed to *make_ipm()* to generate projection kernels, population state time series, and environmental state sequences (if these are define in (6)).", echo = FALSE, dpi = 450, fig.height = 8, fig.width = 8}

library(png)

fig <- readPNG("../Figures/ex_worflow.png")

fig_grob <- rasterGrob(fig, interpolate = TRUE)

plt <- qplot(1:10, 1:10, geom = "blank") +
  annotation_custom(fig_grob,
                    xmin = -Inf,
                    xmax = Inf,
                    ymin = -Inf,
                    ymax = Inf) +
  theme_void()

print(plt)


```

## *Case study 1 - A simple IPM*

One use for IPMs is to evaluate potential performance of invasive species in their new range. Furthermore, decomposing these models into sensitivity and elasticity can help identify effective management strategies (Silvertown, Franco & Menges 1996, Caswell 2001, Baxter et al. 2006, Ellner, Rees & Childs 2016). Bogdan & colleagues (2020) constructed a simple IPM for a Carpobrotus species growing north of Tel Aviv in Israel. The data were collected by drones taking aerial imagery of the population in successive years, and then counting flowers on each plant and generating polygons to estimate sizes and survival from year to year. The model included 4 regressions, and a recruit size distribution. Table 1 provides the mathematical formulae, the corresponding R model formula, and the `ipmr` notation for each one. This version includes a simplified model for growth that models variance around the mean growth rate as constant. This is only for demonstration purposes - users can readily implement non-constant models for the variance of growth (or any other paramter) using `ipmr`. The code also offers an alternative implementation that uses `predict()` methods instead of the mathematical form of the regression model to generate the same output. The final part of the case study provides examples of functions that compute kernel sensitivity and elasticity, the per-generation growth rate, and generation time for the model. 

The `ipmr` version of the R code is meant to closely mirror the mathematical notation. This reduces the amount of required coding expertise - this particular model does not require the user to know how to set up integration meshes, handle eviction, or iterate a model. The first version of the `ipmr` code only requires that the user know the functional form and appropriate transformations of their linear predictors (e.g. inverse logit for a logistic regression). The second version of the code relaxes this requirement, and provides an example using `predict()` methods instead of writing the mathematical form of each regression model. Users who don't want to spend much time thinking about the mathematical form (or have specified a complicated state-fate model) can simply use that format instead.

## *Case study 2 - A general age x size IPM*

General IPMs are useful for modelling species with more complex demography than can be described by a single, continuous state variable. In reality, many species have multiple life stages that are not readily described by a single state variable. For example, a tree species whose demography is best described by diameter at breast height (DBH), may produce seeds that enter a seed bank. Seeds cannot have a value for DBH, but their vigor may decline with age (e.g. Eager et al. 2014, Compagnoni et al. 2016). This requires specifying either a second continuous state variable, or a set of discrete states which individuals may enter and leave over the course of their lifetimes. A simple IPM cannot capture this level of detail. 

We use an age- and size-structured IPM from Ellner, Rees & Childs (2016) to illustrate how to create general IPMs with `ipmr`. The R code provided by the book's authors on their Github repository requires 10 user-specified functions, and a `for` loop nested inside of a call to `with` to work (as well as how to use the functions `with` and `within`). The `ipmr` implementation contains a single user-defined function, and this is only necessary to handle the case of lambs that cannot reproduce immediately after birth. Another key feature this case study demonstrates is the suffix syntax for vital rate and kernel expressions (highlighted in bold in the 'ipmr' column in Table 2). The suffixes appended to each variable name in the `ipmr` formulation correspond to the sub- and/or super-scripts used in the mathematical formulation. `ipmr` internally expands the model expressions and substitutes the range of ages or grouping variables in for the suffixes. This allows users to specify their model in a way that closely mirrors its mathematical notation, and saves them from re-typing model definitions many times or using `for` loops to loop over the range of discrete states. 

# **Discusion of additional applications**

We show above that `ipmr` can handle a variety of model implementations, and how it goes beyond the capabilities of existing software. The common data structure used to represent the model (the `proto_ipm`) provides a concise standardized data structure for representing models in future publications' supplementary materials and could go a long way to enhancing reproducibility in a field not known for making data open access.  Additionally, `ipmr` provides a small number of generic functions for implementing a broad array of IPMs. The package is extensively documented online (https://levisc8.github.io/ipmr/index.html) and in vignettes available within the package. This enables to users to focus on the biology underlying their system of interest, rather than implementation details like eviction, integration meshes, and the model iteration procedure. 

Perhaps more importantly though, the flexibility of `ipmr` raises some intriguing possibilities for the future of structured demography. Recent work has highlighted the power of syntheses that harness many structured population models (Adler et al. 2013, Salguero-Gomez et al. 2016,  Compagnoni et al. 2020). Despite the wide variety of models that are currently published in the IPM literature, `ipmr`'s functional approach is able to reproduce nearly all of them without requiring any raw data at all. The `proto_ipm` data structure does not depend on any raw demographic data to generate a model. A database that stores functional forms and parameters of currently published models could generate `proto_ipm`s, which could then recreate existing models for synthesis using the infrastructure included in this package. This could act as an IPM equivalent of the popular COMPADRE and COMADRE matrix population model databases (Salgo 2015 + 2016). 

# **Limitations**

Currently, `ipmr` can only handle density independent models. Density dependent IPMs are not especially common in the literature at the moment, though that may change as increasingly large data sets become available (e.g. Rose et al. 2005, Adler et al. 2010). Time-lagged models are also not yet implemented. Both of these are currently slated for development, and should be available in the next year or so. Additionally, while `ipmr` does handle the implementation step of IPMs and some very basic analyses (e.g. per-capita growth rates, left and right eigenvectors), this is rarely the end of the analysis for researchers. We are developing a separate  R package to do more complicated analyses with IPMs (e.g. life table response experiments, life history trait analysis). This package is designed to work with `ipmr`'s class structure, and, once stable, will provide a complete tool set for analysis of populations structured by continuously distributed traits.

# Conclusion

We implement a framework for building integral projection models in _R_, and show that it is far more flexible than existing options. It is extensively documented both online and in vignettes and help files accessible from within an R session. We demonstrate that, in addition to implementing user-specified models, this framework could serve as an engine to reproduce published models without requiring any raw data used to implement them. This would allow for far more comprehensive syntheses of published work than what is currently possible. Finally, we discuss the limitations of the package, and our plans to address them in future package versions and separate packages. 

# **Citation list**

1. 1.	Bates D., Maechler M., Bolker B. &  Walker S. (2015). Fitting Linear Mixed-Effects Model Using lme4. Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.

2.	Wood, S.N. (2011) Fast stable restricted maximum likelihood and marginal likelihood estimation of semiparametric generalized linear models. Journal of the Royal Statistical Society (B) 73(1):3-36

3.	Metcalf C. J. E., McMahon S. M., Salguero-Gomez R. & Jongejans E. (2013). IPMpack: an R  package for integral projection models. Methods in Ecology and Evolution. 4(2): 195-200. DOI: 10.1111/2041-210x.12001

4.	Ellner SP, Childs DZ, Rees M. (2016) Data-driven modelling of structured populations: a practical guide to the integral projection model. Basel, Switzerland: Springer International Publishing AGEasterling, Ellner & Dixon 2000

5.	Caswell H. (2001) Matrix population models: construction, analysis, and interpretation, 2nd edn. Sunderland, MA: Sinauer Associates Inc

6.	Ramula, Rees & Buckley (2009). Integral projection models perform better for small demographic data sets than matrix population models: a case study of two perennial herbs. Journal of Applied Ecology 46(5): 1048-1053. DOI: 10.1111/j.1365-2664.2009.01706.x

7.	Compagnoni A., Levin S.C., Childs D.Z., Harpole S., Paniw M., Roemer G., ... Knight T.M. (2020). Short-lived plants have stronger demographic responses to climate. bioRxiv. DOI: https://doi.org/10.1101/2020.06.18.160135

8.	Levin S.C. et al. (unpublished). PADRINO: A database of integral projection models.

9.	Salguero-Gómez R, Jones OR, Archer CA, Buckley YM, Che-Castaldo J, Caswell C, … Vaupel JW (2014) The COMPADRE Plant Matrix Database: an online repository for plant population dynamics. Journal of Ecology 103: 202-218. DOI: 10.1111/1365-2745.12334

10.	Salguero‐Gómez R, Jones OR, Archer CR, Bein C, de Buhr H, Farack C, … Vaupel JW (2016) COMADRE: a global database of animal demography. Journal of Animal Ecology 85: 371-384. doi:10.1111/1365-2656.12482

11.	Henry L. and Wickham H. (2020). rlang: Functions for Base Types and Core R and 'Tidyverse'  Features. R package version 0.4.7. https://CRAN.R-project.org/package=rlang

12. Bogdan A., Levin S.C., Salguero-Gomez R., Knight T.M. (unpublished). Demographic analysis of Israeli Carpobrotus populations: management strategies and future directions. _In prep._

14.	Adler P.B., Salguero-Gomez R., Compagnoni A., Hsu J.S., Ray-Mukherjee J., Mbeau-Ache C. & Franco M. (2014). Functional traits explain variation in plant life history strategies. Proceedings of the National Academy of Sciences 111(2): 740-745. DOI: 10.1073/pnas.1315179111

15.	Jones, O., Scheuerlein, A., Salguero-Gómez, R., Giovanni Camarda C., Schaible R., Casper B. B., … Vapuel J.W. (2014) Diversity of ageing across the tree of life. Nature 505, 169–173 . https://doi.org/10.1038/nature12789

15. Louthan A. & Doak D. (2018) Measurement error of state variables creates substantial bias in results of demographic population models. Ecology 99(10): 2308-2317. DOI: https://doi.org/10.1002/ecy.2455

16. Ellner S.P. & Rees M. (2006). Integral Projection Models for species with complex demography. The American Naturalist 167(3): 410-428.

17. Childs D.Z., Coulson T.N., Pemberton J.M., Clutton-Brock T.H. & Rees M. (2011). Predicting trait values and measuring selection in complex life histories: reproductive allocation decisions in Soay sheep. Ecology Letters 14: 985-992. DOI: 10.1111/j.1461-0248.2011.01657.x

18. Coulson T.N. (2012). Integral projection models, their construction and use in posing hypotheses in ecology. Oikos 121: 1337-1350. DOI: 10.1111/j.1600-0706.2012.00035.x
 
19. Crandall R.M. & Knight T.M. (2017). Role of multiple invasion mechanisms and their interaction in regulating the population dynamics of an exotic tree. Journal of Applied Ecology 55(2):885-894. DOI: 10.1111/1365-2664.13020

20. Adler P.B., Ellner S.P. & Levine J.M. (2010). Coexistance of perennial plants: an embarassment of niches. Ecology Letters 13: 1019-1029. DOI: 10.1111/j.1461-0248.2010.01496.x

21. Baxter P.W.J., McCarthy M.A., Possingham H.P., Menkhorst P.W. & McLean N. (2006). Accounting for management costs in sensitivity analyses of matrix population models. Conservation Biology 20(3): 893-905. DOI: 10.1111/j.1523-1739.2006.00378.x

22. Silvertown, Franco & Menges (1996) Interpretation of Elasticty Matrices as an Aid to the Management of Plant Populations for Conservation. _Con. Biol._ 10(2): 591-597. DOI: 10.1046/j.1523-1739.1996.10020591.x

23. Henry L. & Wickham H. (2020). purrr: Functional Programming Tools. R package version 0.3.4. https://CRAN.R-project.org/package=purrr

24. Bache S.M. & Wickham H. (2014). magrittr: A Forward-Pipe Operator for R. R package version 1.5. https://CRAN.R-project.org/package=magrittr

25. Eager E.A., Rebarber R. & Tenhumberg B. (2014). Globabl asymptotic stability of plant-seed bank models. Journal of Mathematical Biology 69: 1-37. DOI: 10.1007/s00285-013-0689-z

26. Merow C., Dahlgren J.P., Metcalf C.J.E., Childs D.Z., Evans M.E.K., Jongejans E., ... McMahon S.M. (2014). Advancing population ecology with integral projection models: a practical guide. Methods in Ecology and Evolution 5: 99-110. DOI: 10.1111/2041-210X.12146

27. Jongejans E., Shea K., Skarpaas O., Kelly D., & Ellner S.P. (2011). Importance of individual and environmental variation for invasive species spread: a spatial integral projection model. Ecology 92(1): 86-97. DOI: 10.1890/09-2226.1

28. Zuidema P.A., Jongejans E., Chien P.D., During H.J., & Schieving F. (2010). Integral projection models for trees: a new parameterization method and a validation of model output. Journal of Ecology 98: 345-355. DOI: 10.1111/j.1365-2745.2009.01626.x


```{r echo = FALSE}

tab_legend <- "Table 1: Translations between mathematical notation, R's formula notation, and ipmr's notation for the simplified version of Bogdan et al.'s Carpobrotus IPM. The ipmr column contains the expressions used in each kernel's definition. R expressions are not provided for sub-kernels and model iteration procedures because they typically require defining functions separately, and there are many ways to do this step (examples are in the R code for each case study in the appendix). The plogis() function may not be familiar to some users. It is from the 'stats' R package and computes the inverse logit transformation of an expression."

knitr::kable(
  data.frame(

    Math    = c("$\\mu^g = \\alpha^g + \\beta^g * z$",
                "$g(z', z) \\sim Norm(\\mu^g, \\sigma^g)$",
                "$logit^{-1}(s(z)) = \\alpha^s + \\beta^s * z$",
                "$log(f^n(z)) = \\alpha^{f^n} + \\beta^{f^n} * z$",
                "$logit^{-1}(f^p(z)) = \\alpha^{f^p} + \\beta^{f^p} * z$",
                "$f^d(z',z) \\sim Norm(\\mu^{f^d}, \\sigma^{f^d})$",
                "$p^r = \\frac{\\# Recruits (t+1)}{\\# flowers (t)}$",
                "$P = s(z) * g(z',z)$",
                "$F(z',z) = f^p(z) * f^ n(z) * f^d(z', z) * p^r$",
                "$n(z', t+1) = \\int_L^U [P(z',z) + F(z',z)] n(z, t) \\mathrm{dz}$"),
    
    
    R = c("`size_2 ~ size_1, family = gaussian()`",
          "`g = dnorm(z_2, mu_g, sd_g)`",
          "`surv ~ size_1, family = binomial()`",
          "`fec ~ size_1, family = poisson()`",
          "`repr ~ size_1, family = binomial()`",
          "`dnorm(z_2, mu_f_d, sigma_f_d)`",
          "`p_r = n_new_recruits / n_flowers`",
          "",
          "",
          ""),
    
    
    ipmr = c("`mu_g = g_int + g_int + g_slope * z`",
             "`g = dnorm(z_2, mu_g, sd_g)`",
             "`s = plogis(s_int + s_slope * z)`",
             "`f_n = exp(f_int + f_slope * z)`",
             "`f_p = plogis(f_p_int + f_p_slope * z)`",
             "`f_d = dnorm(z_2, f_d_mu, f_d_sigma)`",
             "`p_r = n_new / n_flowers`",
             "`P = s * g`",
             "`F = f_p * f_n * f_d * p_r`",
             "`n_z_t_1 = (P + F) %*% n_z_t`")
  ),
  escape    = FALSE,
  col.names = c("Math Formula", "R Formula", "ipmr"),
  caption   = tab_legend
)

```

Insert into table 2 - won't render in Rmd -> word-table for some reason

$n_0(z', t+1) = \sum\limits_{a=0}^M\int_L^UF_a(z',z)n_a(z,t)\mathrm{dz}$

```{r echo = FALSE}

tab_legend <- "Table 2: Translations between mathematical notation, R's formula notation, and ipmr's notation for Ellner, Rees & Childs (2016) Ovis aries IPM. The ipmr column contains the expressions used in each kernel's definition. R expressions are not provided for sub-kernels and model iteration procedures because they typically require defining functions separately, and there are many ways to do this step (examples are in the R code for each case study in the appendix). ipmr supports a suffix based syntax to avoid repetitively typing out the levels of discrete grouping variables. These are represented as 'a' in the Math column, 'age' in the R formula column, and are highlighted in bold in the ipmr column. The plogis() function may not be familiar to some users. It is from the 'stats' R package and computes the inverse logit transformation of an expression."

knitr::kable(
  data.frame(

    Math    = c(
      "$Logit^{-1}(s(z,a)) = \\alpha^s + \\beta^s_z * z + \\beta^s_a * a$",
      "$g(z', z, a) \\sim Norm(\\mu^g_a, \\sigma^g_a)$",
      "$mu^g(z, a) = \\alpha^g + \\beta^g_z * z + \\beta^g_a * a$",
      "$Logit^{-1}(f^p(z,a) = \\alpha^{f^p} + \\beta^{f^p}_z * z + \\beta^{f^p}_a * a$",
      "$Logit^{-1}(r^p(a) = \\alpha^{r^p} + \\beta^{r^p}_a * a$",
      "$\\mu^r = \\alpha^r + \\beta^r_z * z$",
      "$b(z', z) \\sim Norm(\\mu^r, \\sigma^r)$",
      "$P_a(z',z) = s(z, a) * g(z',z, a)$",
      "$F_a(z',z) = s(z, a) * f^p(z, a) * r^p(a) * b(z', z) / 2$",
      "Insert EQ above table here!",
      "$n_a(z', t + 1) = \\int_L^U P_{a-1}(z',z) n_{a-1}(z,t) \\mathrm{dz}$",
      "$n_{M+1}(z', t + 1) = \\int_L^U[P_M(z',z) n_M(z,t) + P_{M+1}(z',z)n_{M+1}(z, t)]\\mathrm{dz}$"
    ),
    
    
    R = c(
      "`surv ~ size_1 + age, family = binomial()`",
      "`size_2 ~ size_1 + age, family = gaussian()`",
      "`g = dnorm(size_2, mu_g_age, sigma_g)`",
      "`repr ~ size_1 + age, family = binomial()`",
      "`recr ~ age, family = binomial()`",
      "`rc_size_2 ~ size_1, family = gaussian()`",
      "`b = dnorm(size_2, mu_rc_size, sigma_rc_size)`",
      "",
      "",
      "",
      "",
      ""
    ),
    
    ipmr = c(
      "s_**age** = plogis(s_int + s_z * z_1 + s_a * **age**)",
      "g_**age** = dnorm(z_2, mu_g_**age**, sigma_g)",
      "mu_g_**age** = g_int + g_z * z + g_a * **age**",
      "f_p_**age** = plogis(f_p_int + f_p_z * z + f_p_a * **age**)",
      "r_p_**age** = plogis(r_p_int + r_p_a * **age**)",
      "rc_size = dnorm(z_2, mu_rc_size, sigma_rc_size)",
      "mu_rc_size = rc_size_int + rc_size_z * z",
      "P_**age** = s_**age** * g_**age** * d_z",
      "F_**age** =   s_**age** * f_p_**age** * r_p_**age** * rc_size / 2",
      "n_0_t_1 = all_ages(F_**age** %*% n_z_**age**_t, '+')",
      "n_z_**age**_t_1 = P_**age**_minus_1 %*% n_z_**age**_minus_1_t",
      "n_z_**max_age**_t_1 = P_**max_age** %*% n_z_**max_age**_t + P_**max_age**_minus_1 %*% n_z_**max_age**_minus_1_t"
    )
  ),
  escape    = FALSE,
  col.names = c("Math Formula", "R Formula", "ipmr"),
  caption   = tab_legend
)

```
